---
title: "repeatVsRetrievalTimes"
author: "Walter"
date: "June 16, 2018"
output: html_document
---
I was trying to find a way to manipulate retrieval times while keeping the H constant.
I read over Hyman's article in 1953 and in experiment 3:
 In Cond. 3, with four alternatives, the occurrence of a stimulus meant that the probability of its recurring on the next trial was 7/10; the probability of some other alternative occurring was 1/10. 
 
 Sylvan Kornblum wrote about the repeatability changing response times 
 even when H is constant in an article in 1969.  I cited both Hyman and Kornblum 
 on the bottom of this page.
 
So theoretically, we can change the repetition probability for all stimuli while still keeping the overall H value the same.  If the H is maximal, so everything has equal probability, we can change the repetition probability from 0.01 to 0.99 for every
item and still keep H constant.

I calculated the mean retrieval time for each repetition condition and 
correlated repeatability with retrieval time.

```{r}
library(matrixStats)
get_retrieval_time <- function(num_traces,monte_sim_number,rnorm_mean,rnorm_sd) {
  sampled_retrieval_times <- matrix(rnorm(num_traces*monte_sim_number,rnorm_mean,rnorm_sd),
                                    ncol=num_traces,
                                    nrow=monte_sim_number)
  min_retrieval_times <- rowMins(sampled_retrieval_times)
  return(mean(min_retrieval_times))
}
numAlts=26
repeatability=c(0.25,0.5,0.75,0.99)
mean_letter_retrieval_time=c()
amount_of_practice <- 1000
```
The H value for the entire model so far is held constant at 4.77.
However, the repeatability is changing from 0.25 to 0.99.

I basically copied Hyman's procedure:
Everytime the inner loop chooses a new item (1 to 26), the probability of choosing
the same item changes to the repeatability probability.
Then, every other item's probability is the remaining probability divided by
number of items - 1. e.g. 25

```{r}
for(r in 1:length(repeatability))
{  
  
  numAlts=26
  # this creates the set of items to choose from
  sampleSet=seq(1,numAlts,1)
  
 
  
    # this vector will contain all the chosen items during the experiment
    high_entropy_trials=c()
    # high entropy is the frequency distribution that the sample function will use
    high_entropy <- c(rep((1/numAlts),numAlts))
    # choice is which item gets chosen each trial 
    choice=sample(sampleSet,1,prob=high_entropy,replace=T)
    # i store choice so that at every iteration, the chances of choosing
    # the same choice from last trial will be determined by the repeatability
    # i declared at the top of the program
    
    for(i in 1:amount_of_practice)
    {    
      
      #everything other item's chance of being chosen is still equal
      # so Entropy is held constant
      high_entropy=rep((1-repeatability[r])/(numAlts-1),numAlts)
      # the chances of choosing the same item from last trial is
      #controlled by the declared repeatability value
      high_entropy[choice]=repeatability[r]
      #store choice so next trial the repeatability can determine how likely 
      #to choose choice again
      choice <- sample(sampleSet,1,prob=high_entropy,replace=T)
      
      high_entropy_trials=c( high_entropy_trials,choice)
    }
    counts=unique(high_entropy_trials)
    letter_trace_frequencies=c()
    # this loop will determine how many traces per item
    # based on how many times it was chosen and placed in hiEntropyTrials
    for(z in 1:length(counts))
    letter_trace_frequencies=c(letter_trace_frequencies,length(which(high_entropy_trials==counts[z])))
    

    letter_trace_frequencies[letter_trace_frequencies==0] <- 1 # for convenience, we always assume there is 1 trace
    letter_retrieval_times     <- unlist(lapply(letter_trace_frequencies,
                                                function(x) {get_retrieval_time(x,100,500,100)}))
    #mean letter retrieval times stores the mean RT for each repeatability value
    mean_letter_retrieval_time= c( mean_letter_retrieval_time,sum(letter_retrieval_times*letter_trace_frequencies)/sum(letter_trace_frequencies))
  

}
cor(repeatability,mean_letter_retrieval_time)^2
```

Even when H is constant, the correlations are dropping.  The r squared is pretty high:
~0.6.

Hyman, R. (1953). Stimulus information as a determinant of reaction time. Journal of Experimental Psychology, 53, 188–196
Kornblum, S. (1969). Sequential determinants of information processing in serial and discrete choice reaction time. Psychological Review, 76, 113–131.