---
title: "SequentialProbability"
author: "Walter"
date: "June 27, 2018"
output: html_document
---


Goal was to make a program that sampled experiment items based on sequential probability
not based ona frequency distribution
Previously we sampled items for our simulations using freq distribution:
```{r}
#high_entropy <- c(0.4,0.3,0.3)
#choice=sample(sampleSet,1,prob=high_entropy,replace=T)

```
In this example item 1 has 0.4 chance and the others have 0.3 chance of being chosen.  
Everytime the experiment chooses another item it takes based on this sampling distribution.  
For a given frequency distribution, we can manipulate the probability of repetition of 
an item twice.  Instead of choosing from the same frequency distribution
everytime we select a new item, we can create a new freq. distribution. that
makes it more likely to select our previous choice again to increase repeatability.


```{r}
probChart=matrix(c(0.7,0.25,0.25,0.3/2,0.7,0.05,0.3/2,0.05,0.7),ncol=3,nrow=3)
probChart
```
In this example chart, the rows represent the previous choice.  Each row is a frequency
distribution, you can check they all add up to 1.  
The chart tells us the probability of choosing the column's item based on the previous 
choice, the row.  
THe probability of choosing item 2 right after item 1 is 0.15.  
You can also see that the diagonal is all 0.7, which means the chances of choosing same
choice is 0.7, the repeatability.

I want to make a program that can 1) calculate the probability of stimuli pairs
and 2) sample from a given row's probability distribution for a given previous choice.

I will explain how I calculated this probability chart later post.
This program will sample items from a stimuli set based on this chart.
If we add up all the sequential probabilities, item 1  has a total probability
of 0.4 and the other items have a probability of 0.3.  

```{r}
colSums(probChart)
```

This program will sample 3 items with a fixedd repeatability value and
one stimuli will be unequally likely to be selected while the others are equally likely:

```{r}
amount_of_practice=1000

repeatability=0.7
numAlts=3
#prob unequal has to be less than 1-repeatability
probUneq=0.4
probEq=(1-probUneq)/(numAlts-1)

A=matrix(c(1,numAlts-1,(numAlts-2),-((probUneq/probEq)*(numAlts-2))),nrow = 2, ncol=2)
b=c((1-repeatability),(probUneq/probEq)*((1-repeatability)/(numAlts-1)+repeatability)
      -repeatability)
#y will be the probability of the unequal one being selected when previous was equal one
y=solve(A,b)[1]
#x will be the probability of the equal one being selected when previous was equal one
x=solve(A,b)[2]


sampleSet=seq(1,numAlts,1)



# this vector will contain all the chosen items during the experiment
high_entropy_trials=c()
# high entropy is the frequency distribution that the sample function will use
# this time high entropy's 1st index will contain the unequal stimulus
high_entropy <- c(probUneq,rep(probEq,numAlts-1))
# choice is which item gets chosen each trial 
choice=sample(sampleSet,1,prob=high_entropy,replace=T)


for(i in 1:amount_of_practice)
{    
  
  #everything other item's chance of being chosen is still equal
  # so Entropy is held constant
  if(choice==1)
  high_entropy=rep((1-repeatability)/(numAlts-1),numAlts)
  else
  {
    high_entropy=rep(x,numAlts)
    high_entropy[1]=y
  }
  # the chances of choosing the same item from last trial is
  #controlled by the declared repeatability value
  high_entropy[choice]=repeatability
  #store choice so next trial the repeatability can determine how likely 
  #to choose choice again
  choice <- sample(sampleSet,1,prob=high_entropy,replace=T)
  
  high_entropy_trials=c( high_entropy_trials,choice)
}

counts=unique(high_entropy_trials)
letter_trace_frequencies=c()
# this loop will determine how many traces per item
# based on how many times it was chosen and placed in hiEntropyTrials
for(z in 1:length(counts))
  letter_trace_frequencies=c(letter_trace_frequencies,length(which(high_entropy_trials==counts[z])))

counts
letter_trace_frequencies
```
If we look at the counts, we see that one of the items got chosen 400 times while the
others were chosen around 300 times.  
THe next step is to manipulate the probability of unequal item so that the set
is less equiprobable, H drops while we fix the repeatability.

Then we can get the mean retrieval times for each H, while we fix repeatability to see if H affects retrieval times.
