---
title             : "Supplemental materials: Instance theory predicts information theory: Episodic uncertainty as a determinant of keystroke dynamics"
shorttitle        : "Supplemental materials"

author: 
  - name          : "Matthew J. C. Crump"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Brooklyn College of CUNY, 2900 Bedford Avenue, Brooklyn, NY, 11210"
    email         : "mcrump@brooklyn.cuny.edu"
  - name          : "Walter Lai"
    affiliation   : "1"
  - name          : "Nicholaus P. Brosowsky"
    affiliation   : "2"

affiliation:
  - id            : "1"
    institution   : "Brooklyn College of the City University of New York"
  - id            : "2"
    institution   : "The Graduate Center of the City University of New York"


author_note: |
  In addition to these supplemental materials, all of the data and analysis scripts are available at [https://osf.io/bdnqr/](https://osf.io/bdnqr/) [@Crump_Lai_Brosowsky_2018]. This work was supported by a grant from NSF (1353360) to Matthew Crump.
  
abstract: |
  This .pdf contains supplemental materials.
  
keywords          : ""
wordcount         : ""

bibliography      : EntropyTypingRefs.bib

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no

class             : "man"
output            : papaja::apa6_pdf

header-includes:
  - \setlength{\parskip}{0.25cm plus4mm minus3mm}
---

```{r setup, include=FALSE, eval=TRUE,echo=FALSE}
#options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache = FALSE)
```

# Expanded description of Methods

## Participants

400 participants were recruited from Amazon's mechanical turk (restricted to people from the USA, with over 90% completion rate). Data were only analyzed for the 346 participants who successfully completed the task (98 men, 237 women, 11 no response). Additional demographic information is reported in @behmer_crunching_2017. The procedure was approved by the institutional review board at Brooklyn College of the City University of New York. 

## Stimuli and Apparatus

From @behmer_crunching_2017, "Typists copy-typed five normal paragraphs from the Simple English Wiki, a version of the online encyclopedia Wikipedia written in basic English. Four of the paragraphs were from the entry about cats (http://simple.wikipedia. org/wiki/Cat), and one paragraph was from the entry for music (http://simple.wikipedia.org/wiki/Music). Each normal paragraph had an average of 131 words (range 124–137)." In addition to typing English paragraphs, each typist also copy typed a paragraph of 5 letter random strings, and a paragraph of 5 letter strings that approximated the bigram structure of English.

The apparatus was a website displaying a textbox containing a single paragraph. Paragraph text was black, presented in 14 pt, Helvetica font. JavaScript was used to record keystroke timestamps in milliseconds. 

## Design and Procedure

From @behmer_crunching_2017, "Participants were instructed to begin typing with the first letter in the paragraph. Correctly typed letters turned green, and typists could only proceed to the next by typing the current letter correctly. After completing the task, participants were presented with a debriefing, and a form to provide any feedback about the task. The task took around 30 to 45 minutes to complete. Participants who completed the task were paid $1."

# Code for reproducing instance theory simulation of Hyman (1953)

```{r, echo=TRUE}
hyman <- list(
  e1_1 = list(stims = 1,
              probs = 1),
  e1_2 = list(stims = 1:2,
              probs = rep(1/2,2)),
  e1_3 = list(stims = 1:3,
              probs = rep(1/3,3)),
  e1_4 = list(stims = 1:4,
              probs = rep(1/4,4)),
  e1_5 = list(stims = 1:5,
              probs = rep(1/5,5)),
  e1_6 = list(stims = 1:6,
              probs = rep(1/6,6)),
  e1_7 = list(stims = 1:7,
              probs = rep(1/7,7)),
  e1_8 = list(stims = 1:8,
              probs = rep(1/8,8)),
  e2_1 = list(stims = 1:2,
              probs = c(9/10,1/10)),
  e2_2 = list(stims = 1:2,
              probs = c(8/10,2/10)),
  e2_3 = list(stims = 1:4,
              probs = c(13/16,rep(1/16,3))),
  e2_4 = list(stims = 1:6,
              probs = c(15/20,rep(1/20,5))),
  e2_5 = list(stims = 1:4,
              probs = c(4/8,2/8,rep(1/8,2))),
  e2_6 = list(stims = 1:6,
              probs = c(5/10,rep(1/10,5))),
  e2_7 = list(stims = 1:8,
              probs = c(8/16,2/16,rep(1/16,6))),
  e2_8 = list(stims = 1:8,
              probs = c(rep(4/16,2),rep(2/16,2),rep(1/16,4))),
  e3_1 = list(stims = c(11,12,21,22),
              probs = c(2/10,8/10,0,0)),
  e3_2 = list(stims = c(11,12,13,
                        21,22,23,
                        31,32,33),
              probs = c(1/10,8/10,1/10,
                        0,0,0,
                        0,0,0)),
  e3_3 =  list(stims = c(11,12,13,14,
                         21,22,23,24,
                         31,32,33,34),
               probs = c(7/10,1/10,1/10,1/10,
                         0,0,0,0,
                         0,0,0,0,
                         0,0,0,0)),
  e3_4 =  list(stims = c(11,12,13,14,
                         21,22,23,24,
                         31,32,33,34),
               probs = c(1/6,3/6,1/6,1/6,
                         0,0,0,0,
                         0,0,0,0,
                         0,0,0,0)),
  e3_5 =  list(stims = c(11,12,13,14,15,16,17,18,
                         21,22,23,24,25,26,27,28,
                         31,32,33,34,35,36,37,38,
                         41,42,43,44,45,46,47,48,
                         51,52,53,54,55,56,57,58,
                         61,62,63,64,65,66,67,68,
                         71,72,73,74,75,76,77,78,
                         81,82,83,84,85,86,87,88),
               probs = c(1/16,9/16,1/16,1/16,1/16,1/16,1/16,1/16,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0)),
  e3_6 = list(stims = c(11,12,13,
                        21,22,23,
                        31,32,33),
              probs = c(0,1/2,1/2,
                        0,0,0,
                        0,0,0)),
  e3_7 =  list(stims = c(11,12,13,14,15,
                         21,22,23,24,25,
                         31,32,33,34,35,
                         41,42,43,44,45),
               probs = c(0,1/4,1/4,1/4,1/4,
                         0,0,0,0,0,
                         0,0,0,0,0,
                         0,0,0,0,0,
                         0,0,0,0,0)),
  e3_8 =  list(stims = c(11,12,13,14,15,16,17,18,
                         21,22,23,24,25,26,27,28,
                         31,32,33,34,35,36,37,38,
                         41,42,43,44,45,46,47,48,
                         51,52,53,54,55,56,57,58,
                         61,62,63,64,65,66,67,68,
                         71,72,73,74,75,76,77,78,
                         81,82,83,84,85,86,87,88),
               probs = c(0,1/7,1/7,1/7,1/7,1/7,1/7,1/7,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0,
                         0,0,0,0,0,0,0,0))
)

estimate_retrieval_time <-function(n_traces,mean,sd){
  return(mean(replicate(5000,min(rnorm(n_traces,mean,sd)))))
}

sim_df <- data.frame()
for(i in 1:length(hyman)){

  practice <- c(10,50,100,500,1000,5000)
  for(j in 1:length(practice)){
    mean_retrieval_time <-c()
    cnt<-0
    for(k in 1:length(hyman[[i]]$stims)){
      trace_frequency <- round(practice[j]*hyman[[i]]$probs[k])
      if(hyman[[i]]$probs[k] != 0){
        if(trace_frequency < 1){
          trace_frequency <- 1
        }
        cnt<-cnt+1
        mean_retrieval_time[cnt] <- estimate_retrieval_time(trace_frequency,500,100)
      }
    }
    the_probs <-hyman[[i]]$probs
    the_probs <- the_probs[the_probs > 0]
    cond_mean_retrieval_time <- sum(mean_retrieval_time*the_probs)
    t_df <- data.frame(condition = names(hyman[i]),
                       practice = practice[j],
                       bits = sum(hyman[[i]]$probs*log2(hyman[[i]]$probs),
                                  na.rm=TRUE)*-1,
                       modelRT = cond_mean_retrieval_time)
    sim_df <- rbind(sim_df,t_df)
  }
}

library(ggplot2)

sim_df <- cbind(sim_df,
                experiment = rep(c("E1","E2","E3"),each=8*6))

source("ggplot_regressioneqn.R")

figureHyman <- ggplot(sim_df, aes(x=bits,
                                  y=modelRT,
                                  shape=experiment,
                                  color=experiment, 
                                  group=NA))+
  stat_smooth_func(geom="text",method="lm",hjust=0,
                   parse=TRUE,
                   size=2.5, show.legend = FALSE) +
  geom_smooth(method="lm",se=FALSE, color="black", size=.5)+
  geom_point()+
  scale_colour_grey(end=.5)+
  labs(color="Hyman \n (1953)")+
  labs(shape="Hyman \n (1953)")+
  theme_classic()+
  xlab("Bits (H)")+
  ylab("Simulated mean RT")+
  facet_wrap(~practice, scales="free_y")
  

figureHyman

```



# References


\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
